A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2020-01-20 paper.Rmd
update makefile to address #109

suppressMessages(library(ggrepel))
suppressMessages(library(tibble))
suppressMessages(library(kableExtra))
suppressMessages(library(magick))
suppressMessages(library(ggplot2))
suppressMessages(library(cowplot))
```

@@

The README should be placed in the top level of the project, and with one README per dataset. It should be brief, and provide links to the other aforementioned sections, and be in one directory. Saving a README with the extension `.md` file gives the author the formatting benefits of `markdown`, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

@@

```{r variables, fig.cap = "Four tables - the data, and it's variable names, variable codes, and the meaning of missing values", fig.height = 8}
library(tibble)

dat <- tibble::tribble(
  ~job_position, ~outside, ~statistics, ~date_started,
   "Lecturer",        1,        1,      "2019-01-01",
   "Professor",       0,        0,      NA,
   "Senior Lecturer", 1,        1,      "unknown"
)

variable_labels <- tibble::tribble(
  ~variable, ~description,
  "job_position", "University Job Position",
  "outside",      "Did this person study outside this university?",
  "statistics",   "Is this perosn in the statistics faculty",
  "date_started", "What date did this perosn start their position?"
)

variable_codes <- tibble::tribble(
  ~variable,        ~code,
   "job_position",   "text description of job position",
   "outside",       "0 = not studied outside, \n1 = has studied outside",
   "statistics",    "0 = is not in statistics dept., \n1 = is currently in statistics dept.",
   "date_started",   "date variable in format YYYY-MM-DD, \ne.g., 2001-01-01 is 1st January, 2001"
)

missing_values <- tibble::tribble(
  ~value, ~meaning,
  NA,     "Missing value",
  "unknown", "Missing value"
)

library(gridExtra)
grid.arrange(
  tableGrob(dat),
  tableGrob(variable_labels),
  tableGrob(variable_codes),
  tableGrob(missing_values),
  ncol = 1
)

@@

Data with a license clearly establishes rules on how everyone can modify, use, and share data. Without a license, these rules are unclear, and can lead to problems with attribution and citation. It can be overwhelming to try and the right license for a use case. Two licenses that are well suited for data sharing are:

@@

@@

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in repositories like Dryad or Zenodo, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is "minted" by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means.

@@

@@

Raw data is usually the first format of the data provided before any tidying or cleaning of the data. If the raw data is a practical size to share, it should be shared in a folder called `raw-data`. The raw data should be in the form that was first received, even if it is in binary or some proprietary format.  If possible, data dictionaries of the raw data should be provided in this folder as well.

@@

Any code used to clean and tidy the raw data should be provided in the `raw-data` directory. Ideally this would involve only scripted languages, but if other practical steps were taken to clean up the data, these should be recorded in a plain text or markdown file.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small and large. For example, small data could be tabular, as a CSV, in the order of bytes to megabytes to gigabytes that can be easily compressed, and large could be what falls under the umbrella of **big data**. The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as `arkdb` [@arkdb] one could easily chunk large data from flat text files to these lite databases without running into memory limitations.

@@

If the dataset you are shipping has a research application, the most relevant home would be the research data repositories **zenodo** or **dryad**. Zenodo, Launched in 2013 in a joint collaboration between openAIRE and CERN, provides a free, archival location for any researcher to depositor their datasets. The only limits to file sizes are 50gb for individual files, which is generous enough to accommodate a large number of use cases. Zenodo is able to accommodate larger file sizes upon request. The Dryad Digital Repository (@dryad) will take data from any field of research, and perform human quality control and assistance  of the data, with the ability to link data with a journal publication, in exchange for a data publishing fee.

>>>>

suppressMessages(library(ggrepel))
suppressMessages(library(tibble))
suppressMessages(library(kableExtra))
suppressMessages(library(magick))
suppressMessages(library(ggplot2))
suppressMessages(library(cowplot))
suppressMessages(library(gridExtra))
```

@@

The README should be placed in the top level of the project, and with one README per dataset. It should be brief, and provide links to the other aforementioned sections, and be in one directory. It should also contain any other guidance for the user on how to read and interpret the whole directory. For example, explaining where tidy and raw data are stored, and other scripts for tidying.

Saving a README with the extension `.md` file gives the author the formatting benefits of `markdown`, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

@@

```{r variables, fig.cap = "Four tables - the data, and it's variable names, variable codes, and the meaning of missing values", fig.height = 8}

dat <- tribble(
  ~job_position, ~outside, ~statistics, ~date_started,
   "Lecturer",        1,        1,      "2019-01-01",
   "Professor",       0,        0,      NA,
   "Senior Lecturer", 1,        1,      "unknown"
)

variable_labels <- tribble(
  ~variable, ~description,
  "job_position", "University Job Position",
  "outside",      "Did this person study outside this university?",
  "statistics",   "Is this perosn in the statistics faculty",
  "date_started", "What date did this perosn start their position?"
)

variable_codes <- tribble(
  ~variable,        ~code,
   "job_position",   "text description of job position",
   "outside",       "0 = not studied outside, \n1 = has studied outside",
   "statistics",    "0 = is not in statistics dept., \n1 = is currently in statistics dept.",
   "date_started",   "date variable in format YYYY-MM-DD, \ne.g., 2001-01-01 is 1st January, 2001"
)

missing_values <- tribble(
  ~value, ~meaning,
  NA,     "Missing value",
  "unknown", "Missing value"
)

grid.arrange(
  tableGrob(dat),
  tableGrob(variable_labels),
  tableGrob(variable_codes),
  tableGrob(missing_values),
  ncol = 1
)

@@

Data with a license clearly establishes rules on how everyone can modify, use, and share data. Without a license, these rules are unclear, and can lead to problems with attribution and citation. It can be overwhelming to try and find the right license for a use case. Two licenses that are well suited for data sharing are:

@@

Once you decide on a license, you should provide a LICENSE or LICENSE.md file that contains the entire license in the top level of the directory.

@@

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in repositories like Dryad or Zenodo, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is "minted" by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named `citation` should be placed in the directory, at the same level as the `README`. It should contain a DOI, and could be in `.bibtex` format.

@@

To create appropriate metadata, we recommend metadata generators such as `dataspice` or `codebook` [@dataspice; @Arslan2019]. These help automatically create the strucutred JSON data. This metadata should be provided in a folder called "metadata", which should be provided for every dataset.

@@

Raw data is usually the first format of the data provided before any tidying or cleaning of the data. If the raw data is a practical size to share, it should be shared in a folder called `data-raw`. The raw data should be in the form that was first received, even if it is in binary or some proprietary format.  If possible, data dictionaries of the raw data should be provided in this folder as well.

@@

Any code used to clean and tidy the raw data should be provided in the `data-raw` directory. Ideally this would involve only scripted languages, but if other practical steps were taken to clean up the data, these should be recorded in a plain text or markdown file.

@@

Another common situation that researchers face is in dealing with data that fall somewhere between small and large. For example, small data could be tabular, as a CSV, in the order of bytes to megabytes to gigabytes that can be easily compressed, and large could be what falls under the umbrella of **big data**. The happy medium is generally data that are too big to fit on memory of most standard machines, but can successfully fit on disk (https://ropensci.github.io/arkdb/articles/articles/noapi.html). In this case, users who do not have the support to maintain resource intensive database servers can instead rely on light serverless databases such as MonetDB or SQLite. These databases provide disk based storage and using language agnostic interfaces, a analyst can easily query these data in manageable chunks that don't overrun memory limitations. Using software packages such as `arkdb` [@arkdb] one could easily chunk large data from flat text files to these lite databases without running into memory limitations. Another option is to break large text data files into chunks with named sequences (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.).

@@

If the dataset you are shipping has a research application, the most relevant home would be the research data repositories **zenodo** or **dryad**. Zenodo, Launched in 2013 in a joint collaboration between openAIRE and CERN, provides a free, archival location for any researcher to deposit their datasets. The only limits to file sizes are 50gb for individual files, which is generous enough to accommodate a large number of use cases. Zenodo is able to accommodate larger file sizes upon request. The Dryad Digital Repository (@dryad) will take data from any field of research, and perform human quality control and assistance  of the data, with the ability to link data with a journal publication, in exchange for a data publishing fee.
